Modal Verification Patterns for Systems

### 2025-06-23 paper.tex
+

\subsection{Virtualization in Systems Software}
\label{sec:background}
One of the most common forms of abstraction provided by systems software is \emph{virtualization},
which abstracts the relationship between conceptual and physical computing resources. 
Operating System (OS) kernels virtualize memory locations and quantity (via virtual memory and paging~\cite{Denning1970VM}).
Distributed language runtimes may virtualize addresses, even when processes may migrate across machines~\cite{jul1988fine}.
Filesystems virtualize locations on disk~\cite{rodeh2013btrfs,hitz1994file,Rosenblum1992LFS,bonwick2003zettabyte}.
Programs built on top of the corresponding systems software layer work logically at the level of these virtualized resources,
and it makes sense to specify the systems software directly in terms of those abstractions.
\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees, page tables, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification.
\paragraph{Virtualization of Memory Locations}
\label{sec:backgroundonmachinemodel}
\todo{change the text}
\todo[inline,color=pink]{Based on the note I took above, I changed the text}
A typical general purpose computers make the memory resources in RAM as \emph{virtually} available. In other words, a program asking for a memory unit from an operating system kernel is served with a memory address that is \emph{virtualized} such that it may be \emph{mapped to} a physical RAM location or not. As our informal definition suggests, the common technique for \emph{virtualization} of memory locations in OS kernels relies \emph{translating}(\emph{mapping}) virtual addresses into physical ones. The conceptual address-translation map is implemented with processor's(CPU) page-table trees as shown in Figure \ref{fig:pagetables2.png}. In this Figure, we see a typical 4 levels of page tables, a virtual address on the top indexing into the different levels of the page-table tree. Fabrication of virtual addresses out of limited physical addresses is provided with entries in the page-table tree where each points to a physical address aligned to 4KB (4096 byte) boundaries. An address translation requires traversal of a series of tables starting from level 4, and it is conventionally referred as a \emph{page-table walk}. The traversal ends in the level 1 page table with the final lookup to the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.
\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables2.png}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
\end{figure}

\todo[inline,color=cyan]{Ismail, is this text copy-pasted from the VMM paper? 
It has phrasing like ``In this paper...''
That's actually a major issue, self-plagiarism of another peer-reviewed piece of work can get
you into a lot of trouble.
}\todo{inline,color=pink}{I am well-aware what counts as as (including self-plagiarism). These were annotated with todos including cautions for changing the text.}
\paragraph{Virtual Memory Managers with Multiple Address-Spaces}
\label{sec:backgroundonvmm}
\todo{change the text}
The page tables mapping virtual addresses to physical addresses are managed by the OS. Typically each process has its on virtul-to-physical mappings represented by its own page table, which the OS registers with the CPU by storing it in a specific register (\texttt{cr3}) as part of switching to a new process. Using different mappings, which map only disjoint portions of physical memory (with some exceptions in the next section) is how the OS ensures memory isolation between processes.

Most OS kernels have a component called a \emph{Virtual Memory Manager} (VMM)\footnote{Not to be confused with Virtual Machine Monitor. In this paper we focus on non-virtualized scenarios, but hardware virtualization extensions for both x86-64 and ARM make use of an additional set of page tables translating what a \emph{guest} considers to be its (virutalized) physical memory to actual physical memory. Our contributions should offer value in this scenario as well.}
which is responsible for setting up page table mappings and for taking action when a page fault occurs. Often, a page fault indicates a bug in a program --- for example, a null pointer dereference crashes a program not because the hardware designates it an error, but because most OSes refuse to map the first 4K worth of virtual addresses, so dereferencing \texttt{NULL} results in a page fault. In such cases, the OS will terminate the program.

In other cases the OS uses page faults to implement specialized functionality and optimizations. A key example of this is (confusingly) called \emph{paging}: saving room in physical RAM by waiting until memory addresses for code are accessed before copying them to RAM (saving process startup time); or copying memory that has not recently been accessed to disk (i.e., in a \emph{swap} file or partition) and marking those page table entries invalid so a future page fault will give the OS the chance to copy the relevant data back into memory when the program tries to access it again.

The key pieces of functionality a VMM must implement itself, or have available in another OS component (depending on design), are
adding a new page mapping (whether the mapped page contains zeros, file data, or swap data), and removing an existing page mapping.
While this initially sounds like relatively modest functionality whose implementation may be complicated by hardware subtleties, a number of more serious complexities must be handled.
%
%For one, virtual addresses may \emph{alias} when more than one is mapped to the same physical memory. This can occur when a process requests it (e.g., via \texttt{mmap} or \texttt{shm\_open} on UNIX-style kernels). It also occurs automatically in most general-purpose kernels: while for normal programs the OS uses the MMU to isolate and abstract physical memory, the kernel's own functionality is often easier to implement if the kernel can directly access memory given its physical address, for example when interacting with hardware devices. For this reason, most kernels contain an \emph{identity map} of all physical memory in a machine. In some cases this is a literal identity map, where for every virtual address $p$, interpreting $p$ as a virtual address instead will map back to $p$ as a physical address. In other cases there is an offset involved, where to access physical address $p$ the kernel can access virtual address $p+\mathit{offset}$. This identity map saves the kernel the performance cost of having to continually add and remove transient mappings simply to briefly access a specific physical memory region. But this coexists with additional virtual addresses (those used by most kernel code and data structures), in a form of virtual address aliasing. Note that virtual address aliasing is both prevalent in general-purpose kernels, and violates the core assumption of most memory models assumed by verified compilers like CompCert, which assume no virtual address aliasing.
%
%Another complicating factor is that portions of page tables can be shared, i.e., trees of page tables can overlap. One common way this used to happen was for one subtree of page tables to map the kernel if linked in appropriately to a higher-level page table, and then to reuse that kernel mapping portion in every process's page tables. This was mostly phased out in the wake of Spectre
%\mytodo{Colin finish}

The OS scheduler, since it is responsible for dealing with multiple address spaces, must keep track of which virtual addresses are valid (and in what way) in which address spaces, where some virtual addresses are valid in only a single address space (e.g., a code address for a particular usermode process), while others are valid in all address spaces (e.g., kernel data structure pointers). The VMM must maintain some of these assumptions on behalf of the rest of the kernel, for example by guaranteeing that a certain range of virtual addresses (corresponding to the kernel's code and data) are valid in every address space.

\paragraph{Virtualization of Disk-Blocks in Filesystems}
\textsf{ModFS} is a copy-on-write (\textsf{CoW}) filesystem, which as described in the introduction performs updates by copying modified data to \emph{new and unused} locations on disk rather than updating in place, ensuring that in the absence of physical storage corruption, there is \emph{always} a fully consistent filesystem on disk, removing the need for recovery from journals in the case of a crash, which can be very expensive for large volumes.
%\textsf{ModFS} is implemented in \textsc{Go}~\cite{donovan2015go}, using the FUSE~\cite{fuse} interface to run a filesystem in usermode. \textsf{ModFS}'s code is translated to \textsf{Roqc} using the \textsf{Goose}~\cite{chajed2019verifying,chajed2021gojournal} tool, which translates a significant subset of \textsf{Go} to \textsf{Roqc} to permit formal verification of \textsc{Go} code using \textsf{Iris}~\cite{jung2018iris}. \textsf{Goose} generates code in an extension of \textsf{Iris}'s \textsf{HeapLang} with libraries for disk manipulation and \textsc{Go} primitive types (e.g., maps) as shown in Figure \ref{fig:modfslayout}.

>>>>

\subsection{Virtualization in Systems Software}
\label{sec:background}
One of the most common forms of abstraction provided by systems software is \emph{virtualization},
which abstracts the relationship between conceptual and physical computing resources. 
Operating System (OS) kernels virtualize memory locations and quantity (via virtual memory and paging~\cite{Denning1970VM}).
Distributed language runtimes may virtualize addresses, even when processes may migrate across machines~\cite{jul1988fine}.
Filesystems virtualize locations on disk~\cite{rodeh2013btrfs,hitz1994file,Rosenblum1992LFS,bonwick2003zettabyte}.
Programs built on top of the corresponding systems software layer work logically at the level of these virtualized resources,
and it makes sense to specify the systems software directly in terms of those abstractions.
\paragraph{Access by Translation}
Accessing virtualized resources via translation is a common way to virtualize notions of location (e.g.,
virtual memory addresses, inodes or object IDs instead of disk addresses~\cite{bonwick2003zettabyte,hitz1994file,Rosenblum1992LFS,rodeh2013btrfs}).
%Accessing the hardware resources in the presence of sharing means having more resource references (references to the virtualized resources) to access the \emph{mapped} resources that are physical and potentially fewer. 
B-trees, page tables, and related structures both behave like maps, when the corresponding physical resources exist as such
just in a different location.
Control over the lookup process (e.g, handling the case of a missing translation entry) allows for additional flexibility,
such as filling holes in sparse files, or demand paging (both from disk or lazily populating anonymous initially-zeroed mappings).
Although the realization of these maps may differ from system to system based on the context (and sometimes hardware details),
they are semantically --- logically --- partial maps, worth treating as such in verification.
\paragraph{Virtualization of Memory Locations}
\label{sec:backgroundonmachinemodel}
\todo{change the text}
\todo[inline,color=pink]{Based on the note I took above, I changed the text}
A typical general purpose computers make the memory resources in RAM as \emph{virtually} available. In other words, a program asking for a memory unit from an operating system kernel is served with a memory address that is \emph{virtualized} such that it may be \emph{mapped to} a physical RAM location or not. As our informal definition suggests, the common technique for \emph{virtualization} of memory locations in OS kernels relies \emph{translating}(\emph{mapping}) virtual addresses into physical ones. The conceptual address-translation map is implemented with processor's(CPU) page-table trees as shown in Figure \ref{fig:pagetables}. In this Figure, we see a typical 4 levels of page tables, a virtual address on the top indexing into the different levels of the page-table tree. Fabrication of virtual addresses out of limited physical addresses is provided with entries in the page-table tree where each points to a physical address aligned to 4KB (4096 byte) boundaries. An address translation requires traversal of a series of tables starting from level 4, and it is conventionally referred as a \emph{page-table walk}. The traversal ends in the level 1 page table with the final lookup to the actual page of physical memory holding the requested data, and the low-order 12 bits being used to index into this page.
\begin{figure}
    \includegraphics[width=\columnwidth]{pagetables2.png}
    \caption{x86-64 page table lookups.}
    \label{fig:pagetables}
\end{figure}

\todo[inline,color=cyan]{Ismail, is this text copy-pasted from the VMM paper? 
It has phrasing like ``In this paper...''
That's actually a major issue, self-plagiarism of another peer-reviewed piece of work can get
you into a lot of trouble.
}\todo{inline,color=pink}{I am well-aware what counts as as (including self-plagiarism). These were annotated with todos including cautions for changing the text.}
\paragraph{Virtual Memory Managers with Multiple Address-Spaces}
\label{sec:backgroundonvmm}
\todo{change the text}
An intuition for the abstraction that \emph{contains} a page-table tree is called an \emph{address-space} of a process whose virtual addresses to physical addresses. A typical OS manages multiple processes at once . Each process is identified by an \emph{unique root address} of its page-table tree. As part of switching to a new process, i.e., loading the address-space mappings as the current view of the memory, the CPU stores the unique root address of the page-table tree in a specific register (\texttt{cr3}) as shown in Figure \ref{fig:pagetables}. Using different mappings, which map only disjoint portions of physical memory (with some exceptions in the next section) is how the OS ensures memory isolation between processes. Virtual memory managers are the modules in OS kernels handling this kind of tasks. The treat address-spaces as containers of virtual addresses so that they can bookkeep which which virtual addresses are valid (and in what way) in which address spaces.
\paragraph{Virtualization of Disk-Blocks in Filesystems}
\textsf{ModFS} is a copy-on-write (\textsf{CoW}) filesystem, which as described in the introduction performs updates by copying modified data to \emph{new and unused} locations on disk rather than updating in place, ensuring that in the absence of physical storage corruption, there is \emph{always} a fully consistent filesystem on disk, removing the need for recovery from journals in the case of a crash, which can be very expensive for large volumes.
%\textsf{ModFS} is implemented in \textsc{Go}~\cite{donovan2015go}, using the FUSE~\cite{fuse} interface to run a filesystem in usermode. \textsf{ModFS}'s code is translated to \textsf{Roqc} using the \textsf{Goose}~\cite{chajed2019verifying,chajed2021gojournal} tool, which translates a significant subset of \textsf{Go} to \textsf{Roqc} to permit formal verification of \textsc{Go} code using \textsf{Iris}~\cite{jung2018iris}. \textsf{Goose} generates code in an extension of \textsf{Iris}'s \textsf{HeapLang} with libraries for disk manipulation and \textsc{Go} primitive types (e.g., maps) as shown in Figure \ref{fig:modfslayout}.
